A Simple Type Theory for Category Theory

### 2022-05-25 paper/paper.tex
stop abusing the word "simple"

%% Next, Figure~\ref{fig:basic-rules} gives the basic structural rules of
%% judgment theory, which model composition and restrictions in a virtual
%% double category. First, we have varialbes and substitution for types,
%% modeling identity and composition of vertical arrows. Next, we have
%% restriction of a judgment to more specific types, which is given by
%% substitution of types for the judgment's parameter variables. Next, we
%% have rules for constructing an ``empty'' context, which is just a
In typical natural-deduction style, all syntactic forms typed in
context admit an action of substitution. For terms, this is the
ordinary type theoretic substitution. Naturally objects $\alpha:\cat C
\vdash a : \cat D$ can be substituted for object variables $\beta :
\cat D$ in other objects. We can also substitute objects into
\emph{sets}. For instance, if we have a hom set in a category
$\punitinXfromYtoZ {\cat C} \alpha {\alpha'}$, and we can define
objects $a$ and $a'$ in the category $\cat C$, then we can substitute
those in to form the hom set between $a$ and $a'$: $\punitinXfromYtoZ
{\cat C} a {a'}$. This semantically is a ``restriction'' of a
profunctor along two functors, and modeling this operation simply as a
substitution considerably simplifies reasoning using profunctors.
%
Finally we have the action of substitution on natural
transformations. A natural transformation is indexed by set variables
$x:R$, whose sets are each parameterized by a contravariant and a
covariant category variable $\alpha;\beta$. We can think of natural
transformations as \emph{polymorphic} in the categories involved, and
so when we make a transformation substitution, we substitute not just
the set variables, but we also \emph{instantiate} the polymorphic
category variables with objects.

@@

\begin{figure}
  \begin{mathpar}
      \inferrule*[right=Covariant Hom formation]
      {\Gamma \pipe d^+\Xi; \alpha : \cat C \vdash  R \and \Gamma \pipe  d^-\Xi; \alpha : \cat C \vdash P}
      {\Gamma \pipe \Xi \vdash \homrallXYtoZ \alpha R P}

@@

One common source of category theoretic arguments is the duality
principle: since the opposite of any category is also a category, any
statement that quantifies over all categories is equally true of their
opposites. Furthermore in a formal language of categories, any
statement can be dualized to a statement where domains and codomains
are swapped and composition is flipped.
%
A similar duality principle holds in SCT.
%
The basic idea is to simply reverse the orientation of all \emph{set}
expressions by swapping which variable is covariant and which is
contravariant.
%
This has the effect of reversing the order of the context $\Phi$.

We can define this as a simple translation $\cdot^* : SCT \to SCT$
defined on syntactic forms of SCT.

@@

Next, we develop the basics of the model theory for SCT. Because SCT
includes not just a syntax for category theory, but is also indexed by
an ordinary context of dependent type theory $\Gamma$, a model of SCT
is not given by a \emph{single} virtual double category, but instead
by an \emph{indexed family} of VDCs, parameterized by an object of a
category that interprets dependent type theory. We call such a
structure a \emph{VDC-hyperdoctrine}. We then show how to construct a
from any VDC, a VDC-hyperdoctrine indexed by the category of sets,
which has the intuitive reading that any of the VDC judgments $\Gamma
\vdash J$ is interpreted as simply a family of $J$-data parameterized
by elements of the set $\Gamma$.

@@

The principle advantage of DCDs over VDCs is that the presentation of
their algebraic structure is considerably simplified and is
straightforward to describe as a Generalized Algebraic Theory in the
sense of \citet{CartmellGAT}. This is because we don't \emph{require}
that our profunctor contexts are lists, but instead merely the fact
that their concatenation forms a strict category. Then the
associativity/identity equations for a DCD are simply those of a
double category.

@@

Next, we can upgrade any DCD in $\Set$ to a DCD internal to
$\Set^\Set$ by taking the presheaf of objects, vertical arrows, etc as
families of objects, vertical arrows etc indexed by the indexing set
$\Gamma$.
\begin{definition}[Powers of Virtual Equipments]
  Let $\mathcal V$ be a virtual equipment and let $X$ be a set.
  Define a virtual equipment, the ``power'' $\mathcal V^X$ as follows.

  \begin{enumerate}
  \item Objects are functions $X \to \mathcal V_0$.
  \item Vertical morphisms $\mathcal V^X_0(A,B)$ are functions
    $\Pi_{x\in X} V_0(A x,B x)$. With pointwise composition, identity.
  \item Horizontal morphisms $\mathcal V^X_1$ are functions $\Gamma
    \to \mathcal V_1$.
  \item A 2-cell from $\Phi$ to $R$ is a family $\Pi_{x\in X} \mathcal
    V^X_h(\Phi x,R x)$.
  \end{enumerate}
  All operations are simply lifted pointwise.
\end{definition}

>>>>

%% Next, Figure~\ref{fig:basic-rules} gives the basic structural rules of
%% judgment theory, which model composition and restrictions in a virtual
%% double category. First, we have varialbes and substitution for types,
%% modeling identity and composition of vertical arrows. Next, we have
%% restriction of a judgment to more specific types, which is given by
%% substitution of types for the judgment's parameter variables. Next, we
%% have rules for constructing an ``empty'' context, which is just a
In typical natural-deduction style, all syntactic forms typed in
context admit an action of substitution. For terms, this is the
ordinary type theoretic substitution. Naturally objects $\alpha:\cat C
\vdash a : \cat D$ can be substituted for object variables $\beta :
\cat D$ in other objects. We can also substitute objects into
\emph{sets}. For instance, if we have a hom set in a category
$\punitinXfromYtoZ {\cat C} \alpha {\alpha'}$, and we can define
objects $a$ and $a'$ in the category $\cat C$, then we can substitute
those in to form the hom set between $a$ and $a'$: $\punitinXfromYtoZ
{\cat C} a {a'}$. This semantically is a ``restriction'' of a
profunctor along two functors, and modeling this operation as a
substitution considerably simplifies reasoning using profunctors.
%
Finally we have the action of substitution on natural
transformations. A natural transformation is indexed by set variables
$x:R$, whose sets are each parameterized by a contravariant and a
covariant category variable $\alpha;\beta$. We can think of natural
transformations as \emph{polymorphic} in the categories involved, and
so when we make a transformation substitution, we substitute not just
the set variables, but we also \emph{instantiate} the polymorphic
category variables with objects.

@@

\begin{figure}
  % TODO: formalize the one-place substitution
  \begin{mathpar}
      \inferrule*[right=Covariant Hom formation]
      {\Gamma \pipe d^+\Xi; \alpha : \cat C \vdash  R \and \Gamma \pipe  d^-\Xi; \alpha : \cat C \vdash P}
      {\Gamma \pipe \Xi \vdash \homrallXYtoZ \alpha R P}

@@

One common source of category theoretic arguments is the duality
principle: since the opposite of any category is also a category, any
statement that quantifies over all categories is equally true of their
opposites. Furthermore in a formal language of categories, any
statement can be dualized to a statement where domains and codomains
are swapped and composition is flipped.
%
A similar duality principle holds in SCT.
%
The basic idea is to reverse the orientation of all \emph{set}
expressions by swapping which variable is covariant and which is
contravariant.
%
This has the effect of reversing the order of the context $\Phi$.

We can define this as a translation $\cdot^* : SCT \to SCT$ defined on
syntactic forms of SCT.

@@

Next, we develop the basics of the model theory for SCT. Because SCT
includes not just a syntax for category theory, but is also indexed by
an ordinary context of dependent type theory $\Gamma$, a model of SCT
is not given by a \emph{single} virtual double category, but instead
by an \emph{indexed family} of VDCs, parameterized by an object of a
category that interprets dependent type theory. We call such a
structure a \emph{VDC-hyperdoctrine}. We then show how to construct a
from any VDC, a VDC-hyperdoctrine indexed by the category of sets,
which has the intuitive reading that any of the VDC judgments $\Gamma
\vdash J$ is interpreted as a family of $J$-data parameterized
by elements of the set $\Gamma$.

@@

The principle advantage of DCDs over VDCs is that the presentation of
their algebraic structure is considerably simplified and is
straightforward to describe as a Generalized Algebraic Theory in the
sense of \citet{CartmellGAT}. This is because we don't \emph{require}
that our profunctor contexts are lists, but instead merely the fact
that their concatenation forms a strict category. Then the
associativity/identity equations for a DCD are precisely those of a
double category.

@@

Next, we can upgrade any DCD in $\Set$ to a DCD internal to
$\Set^{\Set^o}$ by taking the presheaf of objects, vertical arrows, etc as
families of objects, vertical arrows etc indexed by the indexing set
$\Gamma$ as follows:
\begin{definition}[Powers of Virtual Equipments]
  Let $\mathcal V$ be a virtual equipment and let $X$ be a set.
  Define a virtual equipment, the ``power'' $\mathcal V^X$ as follows.

  \begin{enumerate}
  \item Objects are functions $X \to \mathcal V_0$.
  \item Vertical morphisms $\mathcal V^X_0(A,B)$ are functions
    $\Pi_{x\in X} V_0(A x,B x)$. With pointwise composition, identity.
  \item Horizontal morphisms $\mathcal V^X_1$ are functions $\Gamma
    \to \mathcal V_1$.
  \item A 2-cell from $\Phi$ to $R$ is a family $\Pi_{x\in X} \mathcal
    V^X_h(\Phi x,R x)$.
  \end{enumerate}
  All operations are lifted pointwise.
\end{definition}
